---
title: 'Reproducible Research: Peer Assessment 2'
output:
  pdf_document: default
  html_document:
    keep_md: yes
---

#Events with the most impact on economy and public health in USA

##Synopis
The goal of this report is to explore the US National Oceanic and Atmospheric Administration’s (NOAA) storm database and answer two questions about severe weather events, from the time period 1950 to 2011.

1.Across the United States, which types of events are most harmful with respect to population health?

2.Across the United States, which types of events have the greatest economic consequences?

The results show that during that timeframe, tornados are most harmful with respect to population health, which have caused 5633 deaths and 91346 injuries.

Tornadoes have cost the most property damages costing over 1.5 billion dollars in economic losses.
Excessive wetness have cost the most crop damage costing over 140 million dollars.


##Data processing

The data set was downloaded from the course [page](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) . Description of the dataset can be found [here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) and [here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

The main columns we are interested in :

- EVTYPE : Event type (tornado, blizzard, flood,....)
- FATALITIES : people that died during the natural disaster
- INJURIES : people that got injured  during the natural disaster
- PROPDMG : property damage
- PROPDMGEXP : property daamge exponent
- CROPDMG	: crop damage
- CROPDMGEXP : crop damage exponent

###Load required libraries
```{r}
library(plyr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(car)
```


###Processing the Data
```{r  cache=TRUE}
stormdata <- read.csv("repdata-data-StormData.csv.bz2")
```

Dimensions of the dataset
```{r}
dim(stormdata)
```

Although there are `r ncol(stormdata)` cols, for our analysis we are only interested in the 7 columns mentioned above.

Here is how the raw data looks like for the columns we are interested
```{r}
head(select(stormdata, EVTYPE,FATALITIES,INJURIES,PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP))
```


Info on different event types
```{r}
event_types<-unique(stormdata$EVTYPE)
unique_event<-length(unique(stormdata$EVTYPE))
```

There are `r unique_event` events which is a lot to process. Since we are only interested in events that caused the most financial damange and human casualties, we will limit the events by the resepctive damage.



The crop and property damage has the damage broken down by the base and the exponent.
Below is the distribution of those symbols:

```{r}
table(stormdata$PROPDMGEXP)
```

To clean it, we follow the below stratergy:

- values such as '-','+','?' are mapped to 0
- For the si exponent or the exponent, we map them to the full numeric base......So, k,K,3 becomes 10^3

```{r}
# make 'k', 'K' map to lowercase
stormdata$PROPDMGEXP <- factor(tolower(stormdata$PROPDMGEXP))
stormdata$CROPDMGEXP <- factor(tolower(stormdata$CROPDMGEXP))

# map the base
stormdata$PROPDMGEXP <- as.numeric(recode(as.character(stormdata$PROPDMGEXP), 
    "'0'=1;'1'=10;'2'=10^2;'3'=10^3;'4'=10^4;'5'=10^5;'6'=10^6;'7'=10^7;'8'=10^8;'b'=10^9;'h'=10^2;'k'=10^3;'m'=10^6;'-'=0;'?'=0;'+'=0"))
stormdata$CROPDMGEXP <- as.numeric(recode(as.character(stormdata$CROPDMGEXP), 
    "'0'=1;'1'=10;'2'=10^2;'3'=10^3;'4'=10^4;'5'=10^5;'6'=10^6;'7'=10^7;'8'=10^8;'b'=10^9;'h'=10^2;'k'=10^3;'m'=10^6;'-'=0;'?'=0;'+'=0"))

# calculate value in dollars
stormdata$PROPDMGDOLLAR <- stormdata$PROPDMG * stormdata$PROPDMGEXP
stormdata$CROPDMGDOLLAR <- stormdata$CROPDMG * stormdata$CROPDMGEXP

```

# Results

## Question 1
Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

From the dataset, the columns "FATALITIES" and "INJURIES" which describes human deaths and injuries respectively caused due to the event.


```{r}
evt_fat_injuries<-ddply(stormdata, .(EVTYPE), function(x) colSums(subset(x, select= c(FATALITIES,INJURIES))))

evt_fat_injuries_by_fat<-arrange(evt_fat_injuries,desc(FATALITIES))%>%  select(EVTYPE,FATALITIES) 
evt_fat_injuries_by_inj<-arrange(evt_fat_injuries,desc(INJURIES))%>%  select(EVTYPE,INJURIES) 

```

Fatalities sorted by event in descending order

```{r}
head(evt_fat_injuries_by_fat)
```

Fatalities sorted by event in descending order
```{r}
head(evt_fat_injuries_by_inj)
```

Looking at the data, we see that certain event types cause a lot of fatalities/injuries. So,lets take the minumum casualties we are interested in as greather than 100.

```{r fig.width=15}
#filter data for casualties greaer than 100
#prepare the data for ggplot
evnts_high_damage<- evt_fat_injuries %>% 
  filter(FATALITIES > 100, INJURIES > 100) %>% 
  gather(type,people_affected,FATALITIES:INJURIES) %>% 
  select(Event=EVTYPE,type,people_affected)


ggplot(evnts_high_damage,aes(x = Event,y = people_affected,fill=type)) + 
    geom_bar(stat = "identity") +
    facet_wrap(~type, ncol=1) +
    theme(axis.text.x=element_text(angle = 90)) + 
    ggtitle("Types of Event Causing Injuries/Fatalities Across the US") + 
    labs(y="People affected",fill = "Damage Type")

```

So, we see that tornados cause the most fatalities(5,633) and injuries (91,346) across the USA from 1950 to November 2011.

It is probably not surpising to see there exists a correlation between the number of injuries and deaths with respect to event.


## Question 2
Across the United States, which types of events have the greatest economic consequences?

Economic consequences can be measured with property damage ("PROPDMGDOLLAR") and crop damage ("CROPDMGDOLLAR")

Aggregate property and crop damage
```{r}
evt_damage<-ddply(stormdata, .(EVTYPE), function(x) colSums(subset(x, select= c(PROPDMGDOLLAR,CROPDMGDOLLAR))))

evt_damage_for_prop<-arrange(evt_damage,desc(PROPDMGDOLLAR))%>%  select(EVTYPE,PROPDMGDOLLAR) 
evt_damage_for_crop<-arrange(evt_damage,desc(CROPDMGDOLLAR))%>%  select(EVTYPE,CROPDMGDOLLAR) 

```

Property damage sorted by event in descending order

```{r}
head(evt_damage_for_prop)
```

Crop damage sorted by event in descending order
```{r}
head(evt_damage_for_crop)
```

Top 20 events that cause maximal crop and property damage
``` {r}

top_20_prop_events<-evt_damage_for_prop[1:20,'EVTYPE']
top_20_crop_events<-evt_damage_for_crop[1:20,'EVTYPE']

top_events_damage<-evt_damage[evt_damage$EVTYPE %in% top_20_prop_events | evt_damage$EVTYPE %in% top_20_crop_events , ]
```


```{r fig.height=7, fig.width=15}
# select top events that cause damage
# prepare the data for ggplot
top_events_damage<- top_events_damage %>% 
  gather(type,Damage_in_dollars,PROPDMGDOLLAR:CROPDMGDOLLAR) %>% 
  select(Event=EVTYPE,type,Damage_in_dollars)


ggplot(top_events_damage,aes(x = Event,y = Damage_in_dollars,fill=type)) + 
    geom_bar(stat = "identity") +
    facet_wrap(~type, ncol=1) +
    theme(axis.text.x=element_text(angle = 90)) + 
    ggtitle("Events that cause the most property and  crop damage across the US") + 
    labs(y="Damage in dollars",fill = "Damage Type")

```

From the graph, we see that unlike human fatalities and injuries, there is no correlation between property and event damage with respect to event type.

Tornadoes and wild fires have caused the most property damage: 1.600e+09 and 6.241e+08 respectively.

Excessive wetness and "cold and wet conditions" are the biggest source of crop damges: 1.42e+08 and 6.60e+07 respectively.




